---
title: "Estimating species richness from an aggregation of inventory plots"
author:
  - name: "Eric Marcon"
date: "`r format(Sys.time(), '%d %B %Y')`"
url: https://EricMarcon.github.io/Chao2-Aggregated/
github-repo: EricMarcon/Chao2-Aggregated
# Language
lang: en-US
# Bibliography
bibliography: references.bib
biblio-style: chicago
# LaTeX
preamble: >
  \hyphenation{bio-di-ver-si-ty sap-lings}
# Print table of contents in PDFs?
pdftoc: false
# If true, choose its depth
toc-depth: 3
# Do not modify
always_allow_html: yes
output:
  rmdformats::downcute:
    use_bookdown: yes
    lightbox: yes
  bookdown::pdf_book:
    template: latex/template.tex
    citation_package: natbib
    latex_engine: xelatex
    keep_tex: yes
---

```{r DoNotModify, include=FALSE}
### Utilities. Do not modify.
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos="https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, InstallPackage))
}

# Basic packages
InstallPackages(c("bookdown", "formatR", "kableExtra", "ragg"))

# kableExtra must be loaded 
if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "docx") {
  # Word output (https://stackoverflow.com/questions/35144130/in-knitr-how-can-i-test-for-if-the-output-will-be-pdf-or-word)
  # Do not use autoformat (https://github.com/haozhu233/kableExtra/issues/308)
  options(kableExtra.auto_format = FALSE)
}
library("kableExtra")

# Chunk font size hook: allows size='small' or any valid Latex font size in chunk options
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

```{r Options, include=FALSE}
### Customized options for this document
# Add necessary packages here
Packages <- c("tidyverse", "remotes", "spatstat")
# Install them
InstallPackages(Packages)

# From GitHub
remotes::install_github("EricMarcon/SpatDiv")

# knitr options
knitr::opts_chunk$set(
  cache = TRUE,   # Cache chunk results
  echo = TRUE,     # Show/Hide R chunks
  warning = FALSE, # Show/Hide warnings
  message = FALSE, # Show/Hide messages
  # Figure alignment and size
  fig.align = 'center', out.width = '80%',
  # Graphic devices (ragg_png is better than standard png)
  dev = c("ragg_png", "pdf"),
  # Code chunk format
  tidy = TRUE, tidy.opts = list(blank=FALSE, width.cutoff=50),
  size = "scriptsize", knitr.graphics.auto_pdf = TRUE
  )
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(panel.background=element_rect(fill="transparent", colour=NA),
             plot.background=element_rect(fill="transparent", colour=NA))
knitr::opts_chunk$set(dev.args=list(bg="transparent"))

# Random seed
set.seed(973)
```

# Rationale

The Chao2 estimator [@Chao1987] can be used to estimate the richness of a community from occurrence data in sampling plots.

It has been applied by @CazzollaGatti2022 to abundance data of small sample plots aggregated into occurrence data on a grid.
The question addressed here is the relationship between the aggregation level and the resulting estimation.

The estimation actually does not depend on the aggregation level.

# Data

A large, spatialized community is simulated by package *SpatDiv*.
The community is log-normal.
Spatially, tree species are clustered according to a Thomas process.

Parameters are:

- the size of the square window of the simulation, including the name of units,
- the number of trees per unit area,
- the number of species, 
- the parameters of aggregation: the standard deviation of the displacements of points around the cluster centers (`thomas_scale`) and the average number of trees per cluster (`thomas_mu`).

The simulation may take time and memory over 100,000 trees.

```{r}
library("SpatDiv")
window_size <- 2000
unit_name <- c("meter", "meters")
trees_n_per_area <- 500/10000
species_n <- 500
thomas_scale <- window_size/10
thomas_mu <- 100
```

The community is simulated:

```{r}
library("spatstat")
rSpCommunity(n = 1, 
             size = window_size^2 * trees_n_per_area, 
             S = species_n, 
             Spatial = "Thomas", scale = thomas_scale, mu = thomas_mu,
             win = square(r=window_size, unitname=unit_name)
             ) -> spCommunity
# Number of trees
spCommunity$n
```

Square, random inventory plots are simulated in the community.

Their parameters are:
- their size,
- their number.

```{r}
# Plot side length
side_length <- 30
# Number of plots
plots_n <- 20
```

Random coordinates are drawn:

```{r}
# Draw random plots
X_0 <- runif(plots_n, max = window_size)
Y_0 <- runif(plots_n, max = window_size)
# List of windows
plots_windows <- solapply(seq_along(X_0), function(i) 
  owin(xrange=c(X_0[i], X_0[i] + side_length), yrange=c(Y_0[i], Y_0[i] + side_length), unitname=unit_name))
# Coordinates of the plots in a dataframe
plots_coords <- data.frame(Plot=seq_len(plots_n), X=X_0, Y=Y_0)
```

The plots are mapped:

```{r}
# spatstat
# plot(intersect.owin(spCommunity$window, union.owin(plots_windows)))

# ggplot
window_all <- data.frame(xmin = c(spCommunity$window$xrange[1], pmin(X_0, window_size-side_length)),
                         xmax = c(spCommunity$window$xrange[2], pmin(X_0 + side_length, window_size)),
                         ymin = c(spCommunity$window$yrange[1], pmin(Y_0, window_size-side_length)),
                         ymax = c(spCommunity$window$yrange[2], pmin(Y_0 + side_length, window_size))
                         )
library("ggplot2")
gg_plots <- ggplot() +
  geom_rect(data = window_all, mapping = aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), color="black", fill="transparent") +
  coord_fixed()
print(gg_plots)
```
The sampling intensity is `r format(plots_n * side_length^2 / window_size^2 * 100, digits=3)`% of the total area.

Any plot can be mapped with its content.
The first one is:

```{r}
plot_to_draw <- plots_windows[[1]]
library("dbmss")
autoplot(spCommunity[plot_to_draw])
```



# Inventory

The inventory of a plot is made by the `plot_trees()` function which return a boolean vector for all the trees of the community.
Its value is `TRUE` when a tree is in the plot.

If the plot is partly outside the window, it is moved inside.

```{r}
plot_trees <- function(spCommunity, side_length, x_0, y_0) {
  x_max <- spCommunity$window$xrange[2]
  y_max <- spCommunity$window$yrange[2]

  # The plot must be in the window
  if ((x_0 + side_length) > x_max) x_0 <- x_max - side_length
  if ((y_0 + side_length) > y_max) y_0 <- y_max - side_length

  # Trees in the plot
  is_in <- (spCommunity$x >= x_0) & (spCommunity$x <= x_0 + side_length) &
      (spCommunity$y >= y_0) & (spCommunity$y <= y_0 + side_length)
  return(is_in)
}
```


The inventory is simulated: the trees inside each plot are listed and a abundance by plot dataframe is produced.

```{r}
# Initialize a dataframe with factors
plots_inventory <- data.frame()
# Inventory
for (i in seq_len(plots_n)) {
  plot_inventory <- data.frame(
    Plot = i, 
    Tree = spCommunity$marks$PointType[plot_trees(spCommunity, side_length, X_0[i], Y_0[i])]
    )
  plots_inventory <- rbind(plots_inventory, plot_inventory)
}

# Transform the inventory into an abundance table
library("tidyverse")
plots_inventory %>%
  group_by(Plot, Tree) %>%
  summarise(Abundance = n()) %>%
  pivot_wider(names_from = Tree, values_from = Abundance, values_fill=0) ->
  plots_abundances
```


# Aggregation

plots are grouped in the cells of a grid by the function `group_plots()`.
It returns a three-column dataframe with the plot names and the *x* and *y* coordinates of the left lower corner of the cell of the grid each plot belongs to.

```{r}
# Group plots in a grid
group_plots <- function(plots, grid_x, grid_y) {
  grid_xy <- matrix(0, nrow=nrow(plots), ncol=2)
  for (row in seq_len(nrow(plots))) {
    grid_xy[row, 1] <- grid_x[max(which(plots$X[row] > grid_x))]
    grid_xy[row, 2] <- grid_y[max(which(plots$Y[row] > grid_y))]
  }
  colnames(grid_xy) <- c("x_grid", "y_grid")
  return(grid_xy)
}
```


```{r}
abundance_gridded <- function(abundances, plots, grid_size, x_min=0, x_max=1, y_min=0, y_max=1) {
  
  grid_x <- seq(x_min, x_max, by=grid_size)
  grid_y <- seq(y_min, y_max, by=grid_size)
  plots_gridded <- cbind(plots["Plot"], group_plots(plots, grid_x, grid_y))

  abundances %>%
    inner_join(plots_gridded) %>%
    select(-Plot) %>%
    group_by(x_grid, y_grid) %>%
    summarise_all(sum) ->
    abundances_aggregated
  
  return(abundances_aggregated)
}
```

The `richness_gridded()` function return the estimation of richness by the Chao2 estimator after aggregating the plots of each cell of the grid.

```{r}
richness_gridded <- function(abundances, plots, grid_size, x_min=0, x_max=1, y_min=0, y_max=1) {

  abundances_aggregated <- abundance_gridded(abundances, plots, grid_size, x_min, x_max, y_min, y_max)
  occurences <- apply(abundances_aggregated[, -(1:3)] , 2, function(x) sum(x>0))

  n <- nrow(abundances_aggregated)
  s_obs <- sum(occurences > 0)
  s_1 <- sum(occurences == 1)
  s_2 <- sum(occurences == 2)
  s_chao2 <- ifelse(s_2 > 0,
                    s_obs + (n-1) * s_1^2 / 2 / n / s_2, 
                    s_obs + (n-1) * s_1 * (s_1 -1) / 2 / n 
                    )
  return(s_chao2)
}
```

The estimation of the number of species with a 4-cell grid is:

```{r}
# Example
richness_gridded(plots_abundances,
                 plots_coords,
                 grid_size = window_size/2,
                 x_max = window_size, y_max = window_size)
```
The sensitivity to the grid size is evaluated by dividing it by 2 several times (from 1/2 to 1/32 of the community size).

```{r}
sapply(1:5, function(n) richness_gridded(plots_abundances,
                                         plots_coords,
                                         grid_size = window_size/ 2^n,
                                         x_max = window_size, 
                                         y_max = window_size)
       )
```

The estimation is very stable.

A grid size equal to 1/8th of the window (64 cells) is shown here:

```{r}
grid_interval <- window_size/ 2^3
grid_seq <- seq(grid_interval, window_size * .9999, by=grid_interval)
gg_plots +
  geom_hline(yintercept = grid_seq, col="green") +
  geom_vline(xintercept = grid_seq, col="green")
```

The estimation of richness could be made directly from the abundance data:

```{r}
Richness(colSums(plots_abundances[, -1]))
```


# Turing's relation

```{r}
turing <- function(distribution) {
  n <- sum(distribution)
  s_1 <- sum(distribution == 1)
  cat("Actual number of singletons:", s_1, "\n")
  s_2 <- sum(distribution == 2)
  s_3 <- sum(distribution == 3)
  s_4 <- sum(distribution == 4)
  s_1_hat <- (n-1) /n * 2 *s_2 * (5 *s_2 /6 /s_3 - s_3 /4 /s_4)
  cat("Expected number of singletons:", s_1_hat, "\n")
  return(s_1_hat)
}

# Abundances
turing(colSums(plots_abundances[, -1]))

# Aggregated
abundances_aggregated <- abundance_gridded(plots_abundances,
                                           plots_coords,
                                           grid_size = window_size/ 2^3,
                                           x_max = window_size,
                                           y_max = window_size)
occurences <- apply(abundances_aggregated[, -(1:3)] , 2, function(x) sum(x>0))
turing(occurences)
```




`r if (!knitr:::is_latex_output()) '# References {-}'`
